Detecting confounder and center bias in machine learning models

### 2021-10-26 manuscript.tex

\title{A conditional permutation-based approach to test confounder and center effects in machine learning models}

@@

\begin{abstract} % 150 words
The lack of rigorous non-parametric statistical tests of confounder effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unconfounded} and \emph{fully confounded model}, respectively.
The proposed methods provide a strict control for Type-I errors and high statistical power, even in the presence of typical non-linear and non-normal dependencies between the target variable and the model output.
Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset reveals confoudner effects that were previously unreported and found to be hard to correct for with state-of-the-art confound mitigation approaches.
The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment and improvement of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
\end{abstract}

@@

Predictive modelling and supervised learning methods have recently become increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized}. When evaluating the usefulness and applicability of such markers, predictive performance (i.e. prognostic/diagnostic value) is far from being the only important consideration. Biomedical validity (i.e. whether the model is driven by biomedically relevant signal) and generalizability across contexts and populations are also crucial requirements for candidate biomarkers \citep{woo2017building}.

Spurious, out-of-interest associations between the predictor variables (features) and the prediction target - induced by so-called confounder effects - can be detrimental to the model's biomedical validity and generalizability. Many types of confounder effects can be distinguished.

First, measurement artifacts can obviously be considered as confounders. For instance, in-scanner head motion artifacts in magnetic resonance imaging (MRI)-based predictive models can be especially problematic, if the prediction target is also associated with motor function, like previously demonstrated, among others, in case of Alzheimer's disease \citep{rao2017predictive} attention deficit hyperactivity disorder (ADHD) \citep{couvy2016head, eloyan2012automated} or Autism Spectrum Disorder (ASD) \citep{spisak2014voxel, spisak2019optimal, gotts2013perils}. Likewise, eye-blink artifacts can result in confounder-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.

Second, depending on the research question, demographic and psychometric variables can also be considered as confounders during predictive modelling. For instance, resting-state functional connectivity has been shown to be strongly predictive to age \citep{wang2012decoding, dukart2011age} and moderately to fluid intelligence \citep{he2020deep, cole2012global}. As fluid intelligence is known to decline with aging \citep{kievit2018neural}, models trained to predict intelligence can provide reasonable performance by picking up on solely age-related variance \citep{lohmann2021predicting, dubois2018distributed}.

Third, batch effects or, in multi-center studies, center-effects can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance} as the models can utilize such effects to explain (random or true) center/batch-differences in the predictive target. Such models will display worse-than-expected performance when applied on data from new centers/batches.

Various data cleansing, correction and harmonization approaches have been proposed to mitigate detrimental effects of confounders, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}. However, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest \citep{wachinger2021detect}.

@@

The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via "restricted permutations" that retain the observed ratios of class labels within batches (often referred to as balanced permutations in the literature). However, as shown by \cite{southworth2009properties} and further discussed by \cite{hemerik2018exact}, balanced permutations do not provide a correct null behavior and can be invalid in certain circumstances. Indeed, \cite{ferrari2020measuring} demonstrated that this method does not guarantee a proper type I error control with biased classification problems.
The 'confounder index' (CI) proposed by \cite{ferrari2020measuring}, on the other hand, does not provide any quantification of statistical significance.
Moreover, both approaches involve repeated re-fitting of the machine learning model, which might not be feasible for models with high computational complexity (especially when hyperparameters are optimized in a nested cross-validation).
The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the predictors are a direct cause of the target variable or, alternatively, there is an unobserved random variable that is the cause of both. However, this null hypothesis might not be of interest in many real applications. For instance, causation between several psychological or clinical states and functional MRI activity-connectivity is obviously not \emph{direct}, as functional MRI itself is an \emph{indirect} measure of neural activation. In such cases, the alternative hypothesis of this test might be true in absence of any real confounder.

@@

In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing \citep{dawid1979conditional}. This allows relating the problem to the seminal work of \cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ("no free lunch" theorem). 
As an important consequence, partial correlation - a trivial candidate for investigating the dependence between the predictions and the confounder (given the target variable) - might fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved. These conditions are very common for case of predictive modelling \citep{garcia2009study, kristensen2017whole} (Supplementary Material \ref{sup:nomlinviol})

To overcome the limitations of partial correlation (demonstrated on Fig. \ref{fig:sim-h0-demo}), I build on the permutation-based conditional independence framework of \cite{berrett2020conditional}, to construct two novel tests for evaluating the confounder-bias in predictive models.
The \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder.
The \emph{partial confounder test} investigates whether the model utilizes any confounder-variance, when controlled for the target variable.
Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost (re-fitting the model is not required).
The tests set no assumption about the distribution of the model predictions.
The robustness of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normality and non-linearity).
The proposed tests are then applied on functional brain connectivity data to provide evidence of motion- and center-bias in a model trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} and to evaluate the age and acquisition-batch dependence of fluid intelligence predictions on data from the Human Connectome Project \citep{van2013wu}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}

@@

Obviously, a strong association between $\yhat$ and $\c$ might indicate that the model is biased; its predictions are driven by the confounder rather than information about the target variable.
Testing unconditional independence ($\yhat \independent \c$) between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
For instance, even if $\yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to the dependence of both on $\y$. In other words, if the target variable $\y$ is significantly associated to the confounder variable $\c$, a model that is completely blind to $\c$ (i.e not confounded) might still provide outputs $\yhat$ that are significantly associated with $\c$.

@@

 Conditional independence - in its general form - lies at the heart of several fundamental concepts in statistics and  plays an increasingly important role in various fields of applied statistics, particularly in biomedical applications \citep{spirtes2000causation, peters2016causal, fiedler2011mediation, candes2016panning}. Recently, \cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
 Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional independence testing is effectively impossible. In other words, neither the full nor the partial confounder tests can be constructed so that - for all distributions - they provide a valid type I error control and, at the same time, a non-trivial statistical power.

This result stands in strong contrast to unconditional independence testing - where permutation tests  \citep{pitman1937significance, fisher1942189}, provide a general, distribution-free solution - and it has important implications for confounder testing in predictive modelling where the distribution of the model outputs, depending on the applied machine learning model, is unknown and often heavily non-normal.

One of the trivial candidate for the task, partial correlation, for instance assumes that all involved variables are multivariate Gaussian and - as to be shown below - even its Spearman-based variant is unable to tolerate relatively small deviations from normality and linearity.

@@

Confounder testing was performed on the class probabilities (instead of the predicted labels).

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.png}
  \caption{\textbf{Acquisition- and age-bias of fluid IQ prediction in the HCP dataset.} \\
  Scatter plots and regression lines (with 95\% confidence intervals) show the association of the observed (horizontals axis) and predicted (vertical axis) fluid intelligence scores with various confound regression strategies. Color-coding of the confounder variables (top: acquisition batch, bottom: age group, as shown by the corresponding legends) reveals confounder-bias for both acquisition and age in the models trained on the raw data. This bias is robustly detected by the proposed 'partial' confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
  Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$. Both confound mitigation techniques, but especially COMBAT, improves the predictive performance of the models.
  Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the 'partial confound test'. P-values of the 'full' confound test were all less then 0.0001, i.e. confounder did not fully drive prediction for any of the models.
  }
  \label{fig:hcp}
\end{figure}

Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
The partial confounder test revealed that the model was significantly biased both by age group and acquisition batch (both $p<0.0001$, first column of Fig. \ref{fig:hcp}) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to larger predicted values.

@@

Both feature regression and COMBAT seemed to significantly attenuate the effect of center, however the partial confound test still provided evidence for a significant center-bias ($p=0.04$ and $0.01$ for feature regression and COMBAT, respectively, see the second and third columns of the first row on Fig \ref{fig:abide}).

@@

\end{tabular}
\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination, the corresponding p-values and the p-values of the partial and full confounder tests for all investigated models. Significant confoudner bias is denoted by bold font.}
\end{table}

@@

Confounder-bias can significantly harm the generalizability and biomedical validity of predictive models. Testing confounder-bias of predictive models can be achieved by considering the "confounder-problem" as assessing \emph{conditional independence} between the predicted values and either the target or the confounder variable. These two scenarios result in two different test, testing whether the model is partially or fully driven by the given confounder.
Importantly, in predictive modelling, the model predictions often follow a non-normal distribution, thus any test for confounder bias must be, at least partly, non-parametric.
However, as shown by \citet{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
Here I propose the use of the conditional permutation test (CPT) of \cite{berrett2020conditional} to test confounder bias in predictive models. Adapting this approach for the confounder testing problem makes it possible to place assumptions about the joint distribution of only two out of the three variables involved; that is, it allows constructing both the partial and full confounder tests so that no assumption is made about the distribution of the model output. 
The tests are made available in the python package \emph{mlconfound} which builds upon a parallelized, computationally effective implementation of the Markov-chain Monte-Carlo based parallel pairwise sampler of \cite{berrett2020conditional}. It utilizes coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal. The partial and full confounder tests available in \emph{mlconfound} can be uniformly applied for any predictive model, given that realistic (e.g. cross-validated) model predictions can be obtained. Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost; re-fitting the model is not required.

As expected from theoretical justification, both tests displayed a valid type I error control in the simulations with the necessary assumptions, that is, even if $\yhat$ is not normally distributed. In fact, the tests were found to become somewhat conservative (i.e. lower than expected type I error rates ) in the case of high-performance predictive models (when the $R^2$ between the predicted and observed values is greater than 0.7).

Even tough the tests valid type I error control is justified by theory and reinforced by the simulations, a low statistical power could significantly limit their usefulness. The simulations presented in this paper revealed that both tests have a high, practically relevant power in a wide variety of scenarios; with different strength of the association between the target and the confounder, varying predictive performance, various sample sizes and different true confound-to-signal ratios in the predictions. Considering the partial test, if we apply \citeauthor{cohen2013statistical}'s terminology for the effect size of the confounder bias, sample sizes around 500-1000 can robustly identify a very weak confounder bias (e.g a confounder explaining 1-2\% of the variance in the prediction). A lower sample size of 100 will still very likely identify medium sized confounder-biases (around 9\%). Strong biases  (>25\%) can be very robustly detected with sample sizes as low as 50. 
As expected, the power of the full confounder test is very similar to the partial test, since the only difference is that the conditional contribution of the target to the prediction is investigated, instead that of the confounder.

While different biomedical applications might consider different amounts of bias to be relevant, an a-priory power calculation can help identifying the necessary sample size in order to exclude the possibility of a relevant confounder bias with a high probability.

@@

Among the many potential areas of applications, the proposed tests could be certainly useful in translational neuroscience. A characteristic example is provided by the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds a great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools \citep{woo2017building, wager2013fmri, spisak2020pain}. However, the presence of confounders typical for biomedical research (e.g. sample demographics, center-effects) and specific to the data acquisition and processing approach (e.g. imaging artifacts) presents a great challenge to these efforts and is, currently, often largely neglected.

To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, here they have been deployed in two research settings (a regression and a classification problem) where specific types of confounding effects have previously been described. 
First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
Second, functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g. \cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.

@@

In case of the HCP dataset, the weak but statistically significant association between age and IQ predictions could likely exaggerate to a serious bias when testing the model on data of participants outside of the (relatively narrow)
age range of the HCP sample. In this case, the bias would likely significantly harm the out-of-sample generalizability of this model. The bias of the same model to 'acquisition batch' (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset. There can be manifold reasons for the observed association. Fluid intelligence of the participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g. as a consequence of the human connectome project receiving an increasing degree of public interest during its course, the sociodemographic and educational characteristics of the volunteer sample might have changed over time). To explain the associated change in the target variable, the model might utilise changing characteristics of the imaging data (caused by e.g. slight changes in the calibration settings of the measuring equipment or the changing level of experience of the research staff). This is obviously undesirable for any brain-based predictive model of fluid intelligence.

@@

Motion-bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
Interestingly, in the latter case, the association between the confounder and the predictions become even stronger than before but with an opposite sign. Meanwhile, the prediction performance substantially dropped (although still remained statistically significant). 
A possible explanation to these phenomena is that in-scanner motion, as previously described \citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest (especially in functional connections that are themselves directly associated with motor deficits in ASD).
On the other hand, motion artefacts were previously shown to display complex spatiotemporal patterns in fMRI-based functional connectivity estimates \citep{spisak2014voxel} and simply regressing them out of the features is far from being sufficient for removing their effect on the model. The so-called "residual motion artefacts" and their regional interactions, as described in the referenced work, might be responsible for the paradox behavior of this model, which raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.

>>>>

\title{Detecting confounder and center bias in machine learning models}

@@

%TC:ignore
\begin{abstract} % 150 words
The lack of rigorous non-parametric statistical tests for confounder effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.

Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and, for a given confounder variable, probe the null hypotheses of \emph{unconfounded} and \emph{fully confounded models}, respectively.

The tests provide a strict control for Type-I errors and high statistical power, even for non-linear and non-normal dependencies typical in machine learning.
Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset reveals confounders that were previously partly unreported and found to be hard to correct for with state-of-the-art confound mitigation approaches.

The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment and improvement of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
\end{abstract}
%TC:endignore

@@

Predictive modelling and supervised learning methods have recently become increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized, spisak2020pain, walsh2021dome}. When evaluating the usefulness and applicability of such markers, predictive performance (i.e. prognostic/diagnostic value) is far from being the only important consideration. Biomedical validity (i.e. whether the model is driven by biomedically relevant signal) as well as generalizability and fairness across contexts and populations are crucial requirements for candidate biomarkers \citep{woo2017building, obermeyer2019dissecting, mehrabi2021survey}.

Spurious, out-of-interest associations between the predictor variables (features) and the prediction target - i.e the so-called confounder and collider biases \citep{prosperi2020causal} - can be detrimental to the model's biomedical validity and generalizability. Such biases can originate form many sources.

First, measurement artifacts can obviously be considered as a potential source of confounder bias. For instance, magnetic resonance imaging (MRI)-based predictive models can be driven by in-scanner head motion artifacts, next to (or instead of) signal of neural origin, especially, if the prediction target is also associated with motor function and motion characteristics, like previously demonstrated, among others, in case of Alzheimer's disease \citep{rao2017predictive}, attention deficit hyperactivity disorder (ADHD) \citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD) \citep{gotts2013perils, spisak2014voxel, spisak2019optimal}. Likewise, eye-blink artifacts can result in confounder-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.

Second, depending on the research question, demographic and psychometric variables can also be considered as confounders during predictive modelling. For instance, resting-state functional connectivity has been shown to be strongly predictive to age \citep{dukart2011age, wang2012decoding} and moderately to fluid intelligence \citep{ cole2012global, he2020deep}. As fluid intelligence is known to decline with aging \citep{kievit2018neural}, models trained to predict intelligence might provide a statistically significant predictive performance by picking up solely on age-related variance \citep{dubois2018distributed, lohmann2021predicting}.
Importantly, the target variable must not necessarily display a true association to the confoudner to produce a biased model. Model training can be sensitive to sampling bias and stochastic group differences in the training set; a phenomenon that might have calamitous consequences, as increasingly recognised, for instance, in case of racially biased machine learning models \citep{ obermeyer2019dissecting, lwowski2021risk}.

Third, batch effects or, in multi-center studies, center-effects can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance} as the models can utilize such effects to explain (random or true) center/batch-differences in the predictive target. Such models are hard to calibrate, might output systematically biased predictions and might display a worse-than-expected performance when applied on data from new centers/batches.

Various data cleansing, correction and harmonization approaches have been proposed to mitigate detrimental effects of confounders, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{dukart2011age, spisak2014voxel, abdulkadir2014reduction, rao2017predictive} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}. However, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest \citep{wachinger2021detect}.

@@

The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via 'restricted permutations' that retain the observed ratios of class labels within batches (often referred to as balanced permutations in the literature). However, as shown by \cite{southworth2009properties} and further discussed by \cite{hemerik2018exact}, balanced permutations do not provide a correct null behavior and can be invalid in certain circumstances. Indeed, \cite{ferrari2020measuring} demonstrated that this method does not guarantee a proper type I error control in certain, realistic scenarios.
The 'confounder index' (CI) proposed by \cite{ferrari2020measuring}, on the other hand, does not provide any quantification of statistical significance.
Moreover, both approaches involve repeated re-fitting of the machine learning model, which might not be feasible for models with high a computational cost (especially when hyperparameters are optimized in a nested cross-validation).
The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the predictors are a direct cause of the target variable or, alternatively, there is an unobserved random variable that is the cause of both. However, this null hypothesis might not be of interest in many real applications. For instance, causation between several psychological or clinical states and functional MRI activity/connectivity is obviously not \emph{direct}, as functional MRI itself is an \emph{indirect} measure of neural activation. In such cases, the alternative hypothesis of this test might be true in absence of any real confounder.

@@

In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing \citep{dawid1979conditional}. This allows relating the problem to the seminal work of \cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ('no free lunch' theorem). 
As an important consequence, partial correlation - a trivial candidate for investigating the dependence between the predictions and the confounder (given the target variable) - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (demonstrated on Fig. \ref{fig:sim-h0-demo}). These conditions are very common for case of predictive modelling \citep{garcia2009study, kristensen2017whole} (Supplementary Material \ref{sup:nomlinviol})

To overcome these limitations, I extend the permutation-based conditional independence framework of \cite{berrett2020conditional} with a generalized additive model based conditional distribution estimation, which allows avoiding to place any assumptions about the most problematic variable, the prediction output. 
The framework allows constructing two tests: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost (re-fitting the model is not required).
The robustness of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normality and non-linearity).

The usefulness of the proposed tests in detecting various types of confounder bias is demonstrated in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models. 
First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} is used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{ dubois2018distributed, lohmann2021predicting} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
Second, functional connectivity data from the ABIDE \citep{di2014autism} database is used to investigate the potential motion- and center-bias \cite{gotts2013perils, spisak2014voxel, spisak2019optimal} when training models that aim to predict ASD diagnosis.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%TC:ignore
\section{Methods}

@@

Obviously, a strong association between $\yhat$ and $\c$ might indicate that the model is biased; its predictions are driven by the confounder rather than information about the target variable.
Testing unconditional independence ($\yhat \independent \c$) between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
For instance, even if $\yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to the dependence of both on $\y$. In other words, if the target variable $\y$ displays a true association to the confounder variable $\c$, a model that is completely blind to $\c$ (i.e not confounded at all) might still provide outputs $\yhat$ that are significantly associated with $\c$.

@@

 Conditional independence - in its general form - lies at the heart of several fundamental concepts in statistics and  plays an increasingly important role in various fields of applied statistics, particularly in biomedical applications \citep{spirtes2000causation, fiedler2011mediation, peters2016causal, candes2016panning}. Recently, \cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
 Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional independence testing is effectively impossible. In other words, neither the full nor the partial confounder tests can be constructed so that - for all distributions - they provide a valid type I error control and, at the same time, a non-trivial statistical power.

This result stands in strong contrast to \emph{unconditional} independence testing - where permutation tests  \citep{pitman1937significance, fisher1942189}, provide a general, distribution-free solution - and it has important implications for confounder testing in predictive modelling where the distribution of the model outputs (conditioned on the target variable) - depending on the applied machine learning model - is unknown and often non-normal and non-linear.

One of the trivial candidates for the task, partial correlation, for instance assumes that all involved variables are multivariate Gaussian and - as to be shown below - even its Spearman-based variant is unable to tolerate relatively small deviations from normality and linearity.

@@

Confounder testing was performed on the class probabilities (instead of the predicted labels).
%TC:endignore

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.png}
  \caption{\textbf{Acquisition- and age-bias of fluid IQ prediction in the HCP dataset.} \\
  Scatter plots and regression lines (with 95\% confidence intervals) show the association of the observed (horizontals axis) and predicted (vertical axis) fluid intelligence scores with various confound regression strategies. Color-coding of the confounder variables (top: acquisition batch, bottom: age group, as shown by the corresponding legends) reveals confounder-bias for both acquisition and age in the models trained on the raw data. This bias is robustly detected by the proposed 'partial' confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
  Relation between the observed ($Gf$) and predicted ($\hat{y}$) IQ scores and the confounder variables is shown by the graphs as $R^2 values$. Both confound mitigation techniques, but especially COMBAT, improves the predictive performance of the models.
  Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the 'partial confound test'. P-values of the 'full' confound test were all less then 0.0001, i.e. confounder did not fully drive prediction for any of the models.
  }
  \label{fig:hcp}
\end{figure}

Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $p=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
The partial confounder test revealed that the model was significantly biased both by age group and acquisition batch (both $p<0.0001$, first column of Fig. \ref{fig:hcp}) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to larger predicted values.

@@

Both feature regression and COMBAT seemed to significantly attenuate the effect of imaging center, however the partial confound test still provided evidence for a significant center-bias ($p=0.04$ and $0.01$ for feature regression and COMBAT, respectively, see the second and third columns of the first row on Fig \ref{fig:abide}).

@@

\end{tabular}
\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination, the corresponding p-values and the p-values of the partial and full confounder tests for all investigated confounder-mitigation methods. Significant confoudner bias is denoted by bold font.}
\end{table}

@@

Confounder-bias can significantly harm the generalizability and biomedical validity of predictive models. Testing confounder-bias of predictive models can be achieved by considering the "confounder-problem" as assessing \emph{conditional independence} between the predicted values and either the target or the confounder variable. These two scenarios result in two different tests, probing whether the model is partially or fully driven by the given confounder.
Importantly, in predictive modelling, the model predictions often follow a non-normal distribution, thus any test for confounder bias must be, at least partly, non-parametric.
However, as shown by \citet{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
Here I propose the use of the conditional permutation test (CPT) of \cite{berrett2020conditional} to test confounder bias in predictive models. Adapting this approach for the confounder testing problem allows placing assumptions about the joint distribution of only two out of the three variables involved; that is, it allows constructing both the partial and full confounder tests so that no assumption is made about the distribution of the model output. 
The tests are made available in the python package \emph{mlconfound} which builds upon a parallelized and computationally effective implementation of the Markov-chain Monte-Carlo based parallel pairwise sampler of \cite{berrett2020conditional}. It utilizes coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal. The partial and full confounder tests available in \emph{mlconfound} can be uniformly applied for any predictive model, given that realistic (e.g. cross-validated) model predictions can be obtained. Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost; re-fitting the model is not required.

As expected from theoretical justification, both tests displayed a valid type I error control in the simulations with the necessary assumptions, that is, even if $\yhat$ is not normally distributed. In fact, the tests were found to become more conservative (i.e. lower than expected type I error rates ) in the case of high-performance predictive models (when the $R^2$ between the predicted and observed values is greater than 0.7).

Even though the tests' valid type I error control is justified by theory and reinforced by the simulations, a low statistical power could significantly limit their usefulness. The simulations presented in this paper revealed that both tests have a high, practically relevant power in a wide variety of scenarios. Considering the partial test, if we apply \citeauthor{cohen2013statistical}'s terminology for the effect size of the confounder bias, sample sizes around 500-1000 can robustly identify a very weak confounder bias (e.g a confounder explaining 1-2\% of the variance in the prediction). A lower sample size of 100 will still very likely identify medium sized confounder-biases (around 9\%). Strong biases  (>25\%) can be very robustly detected with sample sizes as low as 50. 
As expected, the power of the full confounder test is very similar to the partial test, since the only difference is that the conditional contribution of the target to the prediction is investigated, instead that of the confounder.

While different biomedical applications might consider different amounts of bias to be relevant, an a-priori power calculation can help identifying the necessary sample size in order to exclude the possibility of a relevant confounder bias with a high probability.

@@

Among the many potential areas of applications, the proposed tests could be particularly useful in translational neuroscience. A characteristic example is provided by the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds a great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making \citep{woo2017building, wager2013fmri, spisak2020pain}. However, the presence of confounders typical for biomedical research (e.g. sample demographics, center-effects) and specific to the data acquisition and processing approach (e.g. imaging artifacts) presents a great challenge to these efforts and is, currently, often largely neglected.

To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, here they have been deployed in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models. 
First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
Second, functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g. \cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.

@@

In case of the HCP dataset, the weak but statistically significant association between age and IQ predictions could likely exaggerate to a serious bias when testing the model on data of participants outside of the (relatively narrow)
age range of the HCP sample. In this case, the bias would likely significantly harm the out-of-sample generalizability of this model. The bias of the same model to 'acquisition batch' (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset. There can be manifold reasons for the observed association. Fluid intelligence of the included participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g. as a consequence of the human connectome project receiving an increasing degree of public interest during its course, the sociodemographic and educational characteristics of the volunteer sample might have changed over time). To explain the associated change in the target variable, the model might utilise changing characteristics of the imaging data (caused by e.g. slight changes in the calibration settings of the measuring equipment or the changing level of experience of the research staff). This is obviously undesirable for any brain-based predictive model of fluid intelligence.

@@

Motion-bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
Interestingly, in the latter case, the association between the confounder and the predictions become even stronger than before but with an opposite sign. Meanwhile, the prediction performance substantially dropped (although still remained statistically significant). 
A possible explanation to these phenomena is that in-scanner motion, as previously described \citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest (especially in functional connections that are themselves directly associated with motor deficits in ASD).
On the other hand, motion artifacts were previously shown to display complex spatiotemporal patterns in fMRI-based functional connectivity estimates \citep{spisak2014voxel} and simply regressing them out of the features is far from being sufficient for removing their effect on the model. The so-called "residual motion artifacts" and their regional interactions, as described in the referenced work, might be responsible for the paradox behavior of this model, which raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.
